---
title: "p8105_hw2_rmk2196.Rmd"
output: github_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1
```{r cars}
library(tidyverse)
library(readxl)

transit_df =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")
  ) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, vending, entrance_type, ada
  ) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```
The data is not tidy. 

```{r}
transit_df |> distinct(line, station_name) |> nrow()

transit_df |> filter(ada == TRUE) |> distinct(line, station_name) |> nrow()

transit_df |> filter(vending == "NO") |> pull(entry) |> mean()
```
1. There are 465 distinct stations
2. There are 84 stations that are ADA compliant
3. 0.3770492 is the proportion of station entrances/exits without vending allow entrance

## Reformat data
```{r}
transit_tidy =
  transit_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)

transit_tidy |> filter(route_name == "A") |> distinct(line, station_name) |> nrow()

transit_tidy |> filter(route_name == "A", ada == TRUE) |> distinct(line, station_name) |> nrow()
```
60 stations serve the A train, and 17 of them are ADA compliant


## Problem 2
```{r}
trash_path = "data/202610 Trash Wheel Collection Data.xlsx"

mr_trash =
  read_excel(trash_path, sheet = "Mr. Trash Wheel", range = cell_cols("A:N")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel"
  )

prof_trash =
  read_excel(
    trash_path, sheet = "Professor Trash Wheel",
    range = cell_cols("A:M"), na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(trash_wheel = "Professor Trash Wheel")

gwynnda =
  read_excel(
    trash_path, sheet = "Gwynnda the Good Wheel of the W",
    range = cell_cols("A:L")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(trash_wheel = "Gwynnda")

trash_df = bind_rows(mr_trash, prof_trash, gwynnda) |> relocate(trash_wheel)
```

The dataset has `r nrow(trash_df)` observations and `r ncol(trash_df)`
variables and each row is one dumpster filled by one of three trash wheels, which are
Mr. Trash Wheel (`r nrow(mr_trash)` dumpsters), Professor Trash Wheel
(`r nrow(prof_trash)`), and Gwynnda (`r nrow(gwynnda)`). Key variables include
`trash_wheel`, which identifies the wheel; `dumpster`, `date`, `month`, and
`year`, which identify when each dumpster was collected; `weight_tons` and
`volume_cubic_yards`, which measure how much trash it held; and counts of
specific items such as `plastic_bottles`, `cigarette_butts`, and `wrappers`.
`homes_powered` estimates how many homes could be powered by the electricity
generated from the trash. Not every wheel records every item: `sports_balls`
is only recorded for Mr. Trash Wheel, and `glass_bottles` is not recorded for
Gwynnda, so these are missing for the other wheels. For the available data,
Professor Trash Wheel collected a total of
`r trash_df |> filter(trash_wheel == "Professor Trash Wheel") |> pull(weight_tons) |> sum(na.rm = TRUE)`
tons of trash, and Gwynnda collected
`r trash_df |> filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE) |> format(big.mark = ",")`
cigarette butts in June 2022.

```{r}
baseline_raw =
  read_csv("data/MCI_baseline.csv", skip = 1, na = c(".", "NA", "")) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = case_match(apoe4, 1 ~ "carrier", 0 ~ "non-carrier")
  )

baseline_df =
  baseline_raw |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)
```
```{r}
nrow(baseline_raw)                                   # recruited
nrow(baseline_df)                                    # met inclusion criteria
baseline_df |> drop_na(age_at_onset) |> nrow()       # developed MCI
baseline_df |> pull(current_age) |> mean()           # average baseline age
baseline_df |> filter(sex == "female") |> pull(apoe4) |> (\(x) mean(x == "carrier"))()
```

```{r}
amyloid_df =
  read_csv("data/mci_amyloid.csv", skip = 1, na = c("NA", "Na", "")) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    values_to = "amyloid_ratio"
  ) |>
  mutate(amyloid_ratio = as.numeric(amyloid_ratio))
```
```{r}
anti_join(baseline_df, amyloid_df, by = "id") |> distinct(id) |> nrow()  # baseline only
anti_join(amyloid_df, baseline_df, by = "id") |> distinct(id) |> nrow()  # amyloid only

mci_df = inner_join(baseline_df, amyloid_df, by = "id")

write_csv(mci_df, "data/mci_combined.csv")
```



